---
title: Overthinking Twilight Imperium, 4th Edition
description: What are the strongest factions? Who has the best heroes?
date: 2025-04-21
categories:
  - Twilight Imperium
  - Stan
format:
  html:
    code-tools: true
    toc: true
execute:
  echo: false
  output: true
  warning: false
  error: false
  freeze: auto
---

_Twilight Imperium_ is a space opera strategy board game by Fantasy Flight Games. The modern board game enthusiasts at Board Game Geek [rank it in their top ten](https://boardgamegeek.com/browse/boardgame?sort=rank&sortdir=asc), but it is also a very complicated game: there are dozens of factions, an expansive technology tree, and almost 50 pages of rules in the [living rules reference](https://images-cdn.fantasyflightgames.com/filer_public/51/55/51552c7f-c05c-445b-84bf-4b073456d008/ti10_pok_living_rules_reference_20_web.pdf).


I found a [data set from thousands of games played online asynchronously](https://lookerstudio.google.com/u/0/reporting/3b435bf2-2100-488c-a424-130f1d22ebb0/page/pE58B) and will use this as a starting point to answer my personal questions about faction and hero strength.

::: {.callout-tip}
_Twilight Imperium_ is a complex strategy game that is, at its heart, about diplomacy and playing the table --- that is, it's a social game (though having the right plastic on the table helps too). Winning the game isn't as simple as just picking the right faction or technology tree.

I do not think data analysis will tell you how to win the game at your table, but I think it is fun to do these kinds of analysis and think the results are an interesting statement on the current "meta" in this version of the game.
:::

## Faction Strength

_Twilight Imperium_ has two dozen different factions, each with their own abilities and flavor of play. I tend to actually really enjoy playing as factions many consider weak, including the Arborec and Mentak Coalition. 


::: {.callout-note collapse="true"}
### Model Math

For each event **E**, we have **P** players (6 in this case) who are playing one of 1-P factions. Each game has one winner drawn from the [categorical distribution](https://mc-stan.org/docs/functions-reference/bounded_discrete_distributions.html#categorical-distribution) based on the factions' abilities.

Here is the code in Stan:

```stan
data {
  int<lower=1> E;                      // number of events
  int<lower=1> P;                      // total number of players
  int<lower=1, upper=P> players[E, 6]; // player IDs per event
  int<lower=1, upper=6> winner[E];     // winner index (1-6)
}
parameters {
  vector[P] ability;                   // faction strength
}
model {
  ability ~ normal(0, 1);             // priors

  for (e in 1:E) {
    vector[6] logit_p;
    for (k in 1:6) {
      logit_p[k] = ability[players[e, k]];
    }
    winner[e] ~ categorical_logit(logit_p);
  }
}
```

:::


```{r}
#| label: setup
library(tidyverse)
library(rstan)
library(reactable)
library(htmltools)
library(htmlwidgets)
library(gt)
library(ggrepel)
library(glmnet)

source(here::here("code/common.R"))

load_or_run_stan_model <- function(
  model_path,
  data,
  file_path,
  ...
) {
  if (file.exists(file_path)) {
    fitted_model <- readRDS(file_path)
    return(fitted_model)
  }
  fitted_model <- rstan::stan(
    file = model_path,
    data = data,
    ...
  )
  saveRDS(fitted_model, file = file_path)
  return(fitted_model)
}
```

```{r}
#| label: load-data

# Load data, selecting 6 player PoK games
# and only taking their final state.
data <- readr::read_delim(
  here::here("drafts/ti4-analysis/data/async_data.csv")
) |>
  select(-`isPOK...13`) |>
  rename(isPOK = `isPOK...6`) |>
  filter(isPOK) |>
  filter(`Player Count` == 6) |>
  filter(factionNameWinner != "Incomplete Match") |>
  filter(!isHomebrewMatch) |>
  filter(roundPlusFinal == "final") |>
  filter(Timestamp >= 1682398800) |>
  mutate(game_id = as.character(setupTimestamp)) |>
  group_by(game_id) |>
  filter(max(score, na.rm = TRUE) == 10) |>
  ungroup()
```

```{r}
#| label: prepare-stan-data-factions
df <- data |>
  mutate(faction_id = as.integer(factor(factionName)))

n_games <- nrow(distinct(df, game_id))

game_positions <- df |>
  distinct(game_id, factionsInGame) |>
  mutate(factionName = str_split(factionsInGame, ", ")) |>
  select(game_id, factionName) |>
  unnest(factionName) |>
  group_by(game_id) |>
  mutate(seat = row_number()) |>
  ungroup()

stan_ready <- df |>
  left_join(game_positions, by = c("game_id", "factionName")) |>
  mutate(win = `Winning Faction` == factionName) |>
  group_by(game_id) |>
  filter(n() == 6) |>
  arrange(game_id, seat) |>
  reframe(
    players = list(faction_id),
    winner = which(win)
  ) |>
  head(n_games)

stan_data <- list(
  E = nrow(stan_ready),
  P = max(df$faction_id),
  players = do.call(rbind, stan_ready$players),
  winner = stan_ready$winner
)
```

```{r}
#| label: fit-faction-model
#| cache: true
fit_base <- load_or_run_stan_model(
  here::here("drafts/ti4-analysis/ti4_20250414.stan"),
  data = stan_data,
  file_path = here::here("drafts/ti4-analysis/data/faction_model.rds"),
  iter = 5000,
  chains = 4,
  cores = 4,
  open_progress = FALSE,
)
```

```{r}
#| label: simulate-factions
#| cache: true
posterior <- rstan::extract(fit_base)
draws <- posterior$ability

n_draws <- nrow(posterior$ability)
n_sim_games <- 1000
faction_levels <- levels(factor(df$factionName))
F <- length(faction_levels)

draw_ids <- sample(1:nrow(draws), n_draws)
posterior_subset <- draws[draw_ids, ]

# Store marginal win prob per draw × faction
marginal_matrix <- matrix(NA_real_, nrow = n_draws, ncol = F)

for (d in 1:n_draws) {
  win_total <- numeric(F)
  appearances <- integer(F)

  for (g in 1:n_sim_games) {
    participants <- sample(1:F, 6)  # 6 unique factions

    scores <- posterior_subset[d, participants]
    probs <- exp(scores) / sum(exp(scores))

    win_total[participants] <- win_total[participants] + probs
    appearances[participants] <- appearances[participants] + 1
  }

  marginal_matrix[d, ] <- win_total / appearances
}
```

```{r}
faction_strength <- tibble(
  faction = faction_levels,
  mean = colMeans(marginal_matrix),
  lower = apply(marginal_matrix, 2, quantile, 0.025),
  upper = apply(marginal_matrix, 2, quantile, 0.975)
)

ggplot(faction_strength, aes(x = reorder(faction, mean), y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  geom_hline(yintercept = 1/6, linetype = "dashed") +
  coord_flip() +
  labs(x = "Faction", y = "Estimated win probability", 
       title = "Marginal Win Probabilities with 95% Credible Intervals")
```

Potential ways to improve this analysis:

+ Account for speaker position and slice strength.

## Faction Combos

The presence of another faction in the game can have a large effect on a faction's strength. For instance, the Yssaril benefit from other factions like the Naalu that have powerful agents they can copy.

```{r}
df <- data |>
  mutate(faction_id = as.integer(factor(factionName)))

n_games <- nrow(distinct(df, game_id))

game_positions <- df |>
  distinct(game_id, factionsInGame) |>
  mutate(factionName = str_split(factionsInGame, ", ")) |>
  select(game_id, factionName) |>
  unnest(factionName) |>
  group_by(game_id) |>
  mutate(seat = row_number()) |>
  ungroup()

stan_ready <- df |>
  left_join(game_positions, by = c("game_id", "factionName")) |>
  mutate(win = `Winning Faction` == factionName) |>
  group_by(game_id) |>
  filter(n() == 6) |>
  arrange(game_id, seat) |>
  reframe(
    players = list(faction_id),
    winner = which(win),
  ) |>
  head(n_games)

stan_data <- list(
  E = nrow(stan_ready),        # Number of games
  P = max(df$faction_id),      # Number of faction types
  players = do.call(rbind, stan_ready$players), # Player ids, PxE
  winner = stan_ready$winner  # Winner index
)
```

```{r}
#| label: fit-pairwise-model
#| cache: true
fit <- load_or_run_stan_model(
  here::here("drafts/ti4-analysis/ti4_pairs.stan"),
  data = stan_data,
  file_path = here::here("drafts/ti4-analysis/data/faction_pairs.rds"),
  iter = 5000,
  chains = 4,
  cores = 4,
  open_progress = FALSE,
)
```

```{r}
#| label: sample-posterior-faction-combos
posterior <- rstan::extract(fit)
interact_draws <- posterior$interact  # This has shape [iterations, P, P]
faction_levels <- levels(factor(df$factionName))
P <- length(faction_levels)
interaction_table <- map_dfr(1:P, function(i) {
  map_dfr(1:P, function(j) {
    tibble(
      from = faction_levels[j],
      to = faction_levels[i],
      effect = interact_draws[, i, j]
    )
  })
})
interaction_summary <- interaction_table %>%
  group_by(from, to) %>%
  summarize(
    mean = mean(effect),
    lower = quantile(effect, 0.025),
    upper = quantile(effect, 0.975),
    .groups = "drop"
  ) %>%
  arrange(desc(mean))
```


```{r}
#| label: tbl-faction-combos
#| column: page-right
#interaction_filtered <- interaction_summary #%>% select(from, to, mean, lower, upper)

faction_names <- unique(sort(interaction_summary$to))
tbl <- reactable(
  interaction_summary,
  filterable = FALSE,
  searchable = FALSE,
  columns = list(
    to = colDef(name = "Target Faction"),
    from = colDef(name = "With Faction"),
    mean = colDef(name = "Effect", format = colFormat(digits = 3)),
    lower = colDef(name = "Lower Bound", format = colFormat(digits = 3)),
    upper = colDef(name = "Upper Bound", format = colFormat(digits = 3))
  ),
  elementId = "faction-interaction-table"
)

# UI + JS Filtering Hook
browsable(
  tagList(
    tags$label("Select Target Faction:"),
    tags$select(
      id = "faction-selector",
      onchange = "filterTable()",
      lapply(faction_names, function(name) tags$option(name))
    ),
    tags$script(HTML(
      "
      function filterTable() {
        var selected = document.getElementById('faction-selector').value;
        var table = Reactable.getInstance('faction-interaction-table');
        if (!table) return;
        table.setFilter('to', selected);
      }
      "
    )),
    onRender(tbl, "
      function(el, x) {
        // Set initial filter to the first option
        var table = Reactable.getInstance('faction-interaction-table');
        if (table) {
          table.setFilter('to', document.getElementById('faction-selector').value);
        }
      }
    ")
  )
)
```

## Heroes

### Most Played Heroes

One of my pet theories is that good heroes are ones that you can play consistently.

```{r}
#| label: tbl-leader-use
#| column: page-right
#| layout-ncol: 2
#| classes: plain
leaders <- data |>
  select(game_id, factionName, leaders) |>
  mutate(leaders = map(leaders, ~ jsonlite::fromJSON(.x))) %>%
  unnest_wider(leaders) |>
  mutate(
    hero_purged = hero == "purged",
    hero_unlocked = hero != "locked"
  )
leaders |>
  group_by(factionName) |>
  summarize(purged = sum(hero_purged)/n()) |>
  arrange(desc(purged)) |>
  gt() |>
  cols_label(
    factionName = "Faction",
    purged = "Hero Purged %"
  ) |>
  fmt_percent(
    columns = c(purged)
  ) |>
  tab_options(
    table.font.name = "inherit",
    column_labels.font.weight = "bold",
    table.margin.right = px(20),
    table.margin.left = px(20)
  ) |>
  gt_split(row_every_n = 14)
```


Here we can see which factions have the most _usable_ heroes [^xxcha]. The Titans hero gives a nice planet upgrade in the home system and gets used in almost every single game. On the other hand, the Barony hero is much more situational.

[^xxcha]: Note that as of Codex III, the Xxcha hero is the only one that _cannot_ be purged.

### Hero Importance

To determine hero strength, we might ask: for each faction, to what extent does playing the hero help them win?

```{r}
#| label: prepare-stan-data-leaders

cannot_purge <- c(
  "Xxcha Kingdom"
)
stan_ready <- df |>
  left_join(game_positions, by = c("game_id", "factionName")) |>
  left_join(leaders, by = c("game_id", "factionName")) |>
  mutate(win = `Winning Faction` == factionName) |>
  mutate(can_purge = !(factionName %in% cannot_purge)) |>
  group_by(game_id) |>
  filter(n() == 6) |>
  arrange(game_id, seat) |>
  reframe(
    players = list(faction_id),
    winner = which(win),
    hero_purged = list(hero_purged),
    hero_unlocked = list(hero_unlocked),
    can_purge = list(can_purge)
  ) |>
  head(n_games)

stan_data <- list(
  E = nrow(stan_ready),
  P = max(df$faction_id),
  players = do.call(rbind, stan_ready$players),
  hero_purged = do.call(rbind, stan_ready$hero_purged),
  hero_unlocked = do.call(rbind, stan_ready$hero_unlocked),
  can_purge = do.call(rbind, stan_ready$can_purge),
  winner = stan_ready$winner
)
```

```{r}
#| label: fit-hero-model

fit <- load_or_run_stan_model(
  here::here("notes/ti4-analysis/ti4_hero.stan"),
  data = stan_data,
  file_path = here::here("notes/ti4-analysis/data/hero_fit.rds"),
  iter = 1000,
  chains = 4,
  cores = 4,
  open_progress = FALSE,
)
```

```{r}
#| label: fig-hero-effects
#| fig-cap: Estimated effect of purging each faction’s hero on win probability in Twilight Imperium. Points show posterior means; error bars show 95% credible intervals. Values above zero indicate factions that tend to perform better when their hero is purged.
posterior <- rstan::extract(fit)
draws_hero <- posterior$hero_effect

hero_effect <- tibble(
  faction = faction_levels,
  mean = colMeans(draws_hero),
  lower = apply(draws_hero, 2, quantile, 0.025),
  upper = apply(draws_hero, 2, quantile, 0.975)
)

ggplot(hero_effect, aes(x = reorder(faction, mean), y = mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(
    x = "Faction",
    y = "Hero Effect on Win Log-Odds",
    title = "Importance of Hero by Faction",
  )
```

Unsurprisingly, the Winnu game seems most dependent on their ability to use their hero, which can let them play the primary ability imperial strategy card at any point. 
